---
title: "Integration with bhmbasket"
author: "Kevin Kunzmann"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bhmbasket-integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include=FALSE}
# see whether jags can be found, if not deactivate execution
jags_found <- tryCatch(
  {
    rjags::jags.version()
    TRUE
  },
  error = function(e) FALSE
)

knitr::opts_chunk$set(
  collapse = TRUE,
  eval = jags_found,
  comment = "#>",
  fig.width = 7,
  fig.height = 7 / 1.61,
  fig.align = "center"
)
```

```{r setup}
library(oncomsm)
library(dplyr, warn.conflicts = FALSE)
library(future) # parallel processing
plan(multisession)
```


**tl;dr:** *The [`bhmbasket`](https://cran.r-project.org/package=bhmbasket) 
package implements Bayesian hierarchical methods for basket trials with binary endpoints.
Although `oncomsm` does not support hierarchical modelling,
`bhmbasket` can be used to define 'go' criteria based on hierarchical models.*

Start by defining a prior specification for a multi-group trial.

```{r prior-specification}
mdl <- create_srpmodel(
  A = define_srp_prior(
      p_mean = 0.3, p_n = 20,
      median_t_q05 = c(3, 2, 6) - 1,
      median_t_q95 = c(3, 2, 6) + 1,
      recruitment_rate = 2
    ),
  B = define_srp_prior(
      p_mean = 0.4, p_n = 20,
      median_t_q05 = c(2.5, 8, 18) - 1,
      median_t_q95 = c(2.5, 8, 18) + 1,
      recruitment_rate = 2
    ),
  C = define_srp_prior(
      p_mean = 0.5, p_n = 20,
      median_t_q05 = c(2, 12, 24) - 1,
      median_t_q95 = c(2, 12, 24) + 1,
      recruitment_rate = 2
    )
)
```


## Defining 'go' decision

The `bhmbasket` package implements dynamic borrowing between arms in
basket trials via Bayesian hierarchical models (BHMs).
Here, we demonstrate how `oncomsm` and the prior specified in `mdl` can be used
to derive probability of 'go' based on a `bhmbasket` analysis. 
We use the "Berry" type model for analysis of the response data and use a
posterior $0.25$ quantile of the response probability above $0.3$ to declare
'go'.
This means that an individual arm is further developed if there is a $0.75$
posterior probability according to the BHM that the response rate is larger 
than $0.3$.

```{r}
go <- function(model, data, nsim = 250) {
  set.seed(2340239L)
  # convert data to bhmbasket format (multi-state to binary counts)
  data <- data %>%
    group_by(group_id, subject_id) %>%
    summarize(
      responder = any(state == "response"),
      .groups = "drop_last"
    ) %>%
    summarize(
      r = sum(responder),
      n = n(),
    ) %>%
    {
      bhmbasket::createTrial(.$n, .$r)
    }
  # define Berry model in bhmbasket
  prms_berry <- bhmbasket::setPriorParametersBerry(
    mu_mean   = bhmbasket::logit(0.25),
    mu_sd     = 1,
    tau_scale = 1
  )
  # perform analysis in bhmbasket
  res <- suppressMessages(bhmbasket::performAnalyses(
    scenario_list         = data,
    method_names          = "berry",
    evidence_levels       = 1 - 0.25,
    prior_parameters_list = prms_berry,
    target_rates          = c(.2, .3, .3),
    n_mcmc_iterations     = nsim,
    verbose               = FALSE
  ))
  # 'go' if posterior quantile of response rate is sufficiently large
  return(tibble(
    group_id = model$group_id,
    go = res$scenario_1$quantiles_list$berry[[1]][5, 2:4] >= .3
  ))
}
```


## Probability of 'go' before start of the trial

The 'go' criterion can then be applied to each of the resampled data sets.

```{r, warning=FALSE}
tbl_decisions <- simulate_decision_rule(mdl,
                                        c(40, 40, 40),
                                        go,
                                        nsim = 250,
                                        seed = 32487)

tbl_pr_go_planning <- tbl_decisions %>%
  group_by(group_id) %>%
  summarize(`Pr[go] planning` = mean(go))

tbl_pr_go_planning
```


## Update probability of 'go'

Now assume that some interim data is available.
The data is fairly extreme for both group A and group B.

```{r}
tbl_interim <- tribble(
   ~subject_id, ~group_id, ~t, ~state,
          "s1", "A", 0, "stable",
          "s1", "A", 1.5, "stable",
          "s1", "A", 2.25, "response",
          "s2", "A", 1, "stable",
          "s2", "A", 2, "response",
          "s3", "A", 3, "stable",
          "s3", "A", 4.5, "response",
          "s4", "B", 0, "stable",
          "s5", "B", 1.5, "stable",
          "s6", "C", 0, "stable",
          "s6", "C", 1.5, "progression",
          "s7", "C", 2.25, "stable",
          "s7", "C", 3, "progression",
          "s8", "C", 2, "stable",
          "s8", "C", 3, "progression",
          "s9", "C", 2, "stable",
          "s9", "C", 2.6, "stable",
          "s9", "C", 3, "progression",
         "s10", "C", 3, "stable",
         "s10", "C", 5, "progression"
)

# plot it
visits_to_mstate(tbl_interim, mdl) %>%
  plot_mstate(mdl, relative_to_sot = FALSE)
```

The prior can now be updated with this data.

```{r}
smpl_prior <- sample_prior(mdl, seed = 2314513)
smpl_posterior <- sample_posterior(mdl, tbl_interim, seed = 2314)

tibble(
  group_id = mdl$group_id,
  prior = rstan::extract(smpl_prior, "p")[[1]] %>% colMeans(),
  posterior = rstan::extract(smpl_posterior, "p")[[1]] %>% colMeans()
)
```

The probability of 'go' is then updated by sampling forward from the 
posterior predictive.

```{r, warning=FALSE}
tbl_decisions_interim <- simulate_decision_rule(
  mdl,
  c(40, 40, 40),
  go,
  data = tbl_interim,
  nsim = 250,
  seed = 32487
)

tbl_pr_go_planning %>%
  left_join(
    tbl_decisions_interim %>%
    group_by(group_id) %>%
    summarize(
      `Pr[go] interim` = mean(go)
    ),
    by = "group_id"
  )
```
