---
title: "Avoiding Bias During Interim Analyses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{avoiding-bias}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7 / 1.61,
  fig.align = "center"
)
```

```{r setup}
library(dplyr)
library(tidyr)
library(oncomsm)
```



**tl;dr:** *During interim looks, individuals might still be stable and thus 
'at risk' of either progression or response. Especially when the respective
follow-up times differ, or if there is a difference in median time to response
and median time to progression, this might lead to bias in the analysis of the
response probability.*

```{r}
mdl <- create_srpmodel(
  A = define_srp_prior(
    median_t_q05 = c(1, 4, 12), # shorter time to response than progression
    median_t_q95 = c(6, 8, 36),
    shape_q05 = c(.99, .99, .99), # fixed shape
    shape_q95 = c(1, 1, 1),
  )
)

plot(mdl, confidence = .9)
```

```{r}
tbl_data_interim <- tibble::tribble(
    ~subject_id, ~t,   ~state,
        "subj1",  1, "stable",
        "subj1",  5, "stable",
        "subj2",  1, "stable",
        "subj2",  5, "stable",
        "subj3",  1, "stable",
        "subj3",  5, "stable",
        "subj4",  1, "stable",
        "subj4",  5, "stable",
        "subj5",  0, "stable",
        "subj5",  1, "response",
        "subj5",  2, "EOF",
        "subj6",  0, "stable",
        "subj6",  6, "progression"
  ) %>% 
  mutate(group_id = "A")
plot_mstate(visits_to_mstate(tbl_data_interim, mdl), mdl, relative_to_sot = FALSE)
``` 

A naive estimate of the response rate at month 6 would be $1/2$.
This is implicitly treating the missing response status of the four individuals
at risk as missing completely at random (MCAR) in a binomial model.
However, given the prior assumption that responses are observed earlier than
progressions (also consistent with observed data),
it is less likely that the four individuals at risk still become responders -
their conditional risk of progression depends on the posterior parameter
estimates of the multi-state model and the duration of their 'stable' state.
Hence, the posterior for the response probability should be concentrated at
a value smaller than $0.5$.

```{r}
p_posterior <- parameter_sample_to_tibble(
    mdl,
    sample_posterior(mdl, tbl_data_interim)
  ) %>% 
  filter(parameter == "p") %>% 
  pull(value)

hist(p_posterior)

summary(p_posterior)
```

Indeed this is the case. 
The multi-state model can thus support a more principled approach to estimating
key quantities like response probability or duration of response during interim
analyses.
